\documentclass[10pt,twocolumn]{article}

% ---------------------------------------------------------------------------
%  Waggle: Name Semantics for Agent Handoffs
%  A systems paper. Classic two-column, Computer Modern, hand-drawn TikZ.
% ---------------------------------------------------------------------------

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[utf8]{inputenc}
\usepackage[hmargin=1.85cm,vmargin=2.1cm,columnsep=0.7cm]{geometry}
\usepackage{microtype}
\usepackage{booktabs}
\usepackage{array}
\usepackage{amsmath,amssymb}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc,fit,backgrounds,shapes.geometric,decorations.pathreplacing}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{caption}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{balance}

\captionsetup{font=small,labelfont=bf}
\setlist{nosep,leftmargin=1.2em}
\SetKwComment{Comment}{$\triangleright$\ }{}
\newcommand{\tok}[1]{\texttt{#1}}
\renewcommand{\ttdefault}{cmtt}

% Consistent muted palette (prints legibly in greyscale).
\definecolor{ink}{HTML}{1A1A1A}
\definecolor{stone}{HTML}{6B6B6B}
\definecolor{pale}{HTML}{EDEDE8}
\definecolor{line}{HTML}{4A4A4A}

\title{\vspace{-1.2em}\textbf{The Dance and the Field:\\[2pt]
Name Semantics for Handoffs Between Distributed Agents}}

\author{
  Chetan Conikee\\
  \normalsize Modiqo\\
  \normalsize\texttt{chetan@modiqo.ai}
}
\date{}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Agent harnesses are becoming distributed systems: a single task now fans
out across dozens of subagents, within one process, across the harnesses
on a machine, and increasingly over the network. Yet the mechanism these
systems use to pass work between agents is the one that distributed
systems abandoned decades ago---\emph{copy the bytes and hope}. A report
produced by one agent is pasted into the context of the next, re-billed
on every subsequent turn of a stateless API, duplicated across every
recipient, and stripped of identity, version, and any record of whether
it was read. Copies are the root defect: a copy has no owner, no
lifecycle, and no receipt, and roughly a third of multi-agent failures
trace to agents acting on divergent copies of what should have been the
same information.

We present \textsc{waggle}, a substrate built on \emph{name semantics}.
A handoff is a $\sim$30-byte attributed token, not a payload. Consumers
do not receive content; they \emph{resolve} the token into a projection
computed for their own context, \emph{interrogate} it under byte budgets
where the bytes live, and leave \emph{receipts} in an append-only,
payload-free log. The same five verbs operate unchanged at three
radii---inter-process, inter-harness, and inter-machine---which is the
paper's central systems argument: portability of a handoff cannot be an
optimization applied to copies; it must be a property of the reference.
We ground the design in a lineage the agent literature has overlooked:
tuple spaces, named-data networking, capabilities, leases, and
stigmergy. We describe a sealed context-adaptive matcher, a
consumption-contract mechanism that turns ``did the subagent actually
read it?'' from an unanswerable question into a query, and a mint-time
structural lens for source code that gives an agent a symbol-addressable
handoff without any parser on the serving path. We report microbenchmarks
(39\,ns cache-hit resolution, 323\,\textmu{}s end-to-end over a domain
socket, a million-event fold in 334\,\textmu{}s) and an evaluation over twelve
artifact shapes---text, markdown, code, PDF, voice, video, a folder of many
files, a $>$2000-line source, a multi-hop chain, a reasoning task no single line
answers, and a 180-file tree interrogated for a needle and a count---in which nine models answer the same question
under four turn-matched, \emph{paired} handoffs (1704 runs) against a steelmanned
baseline---the path arm may \texttt{ls}, \texttt{grep}, and run \texttt{pdftotext}.
A raw path trails the token even where it can read the bytes---on a PDF it reaches
78\,\% with \texttt{pdftotext} against the token's 100\,\% at a fourteenth of the
bytes---and collapses where structure must be \emph{traversed} rather than described
(29\,\% on the chain); against it the token wins 34 paired comparisons and loses 9.
Against \emph{pasting} the token neither wins nor loses: 96.5\,\% against 96.7\,\%, a
statistical tie---bought at \textbf{32$\times$ fewer ingested bytes} and
10$\times$ less money. Pasting was never inaccurate; it was expensive, and it
left no record. It does collapse once, and instructively: asked to \emph{count}
across a 180-file tree, a copy holding every byte answers 61\,\% where the token
answers 100\,\%---context was never the missing thing. Two results outlast the
arms. The receipt is an \emph{oracle}: runs whose coverage showed the required
regions consumed were 99\,\% correct, and those showing them unconsumed 20\,\%,
a seventy-nine-point collapse visible to an orchestrator that never read the answer.
And the benchmark's first service was to find seven defects---in the substrate,
not the models---which taught us to stop using a language model as a bug
detector and write the deterministic conformance suite that should have existed
first. We are candid about the mechanism's limits, including one it made worse.
\end{abstract}

% ===========================================================================
\section{Introduction}
% ===========================================================================

Begin with the fact that shapes everything else: large-language-model
APIs are \emph{stateless}. Each turn of a conversation re-sends the entire
conversation. When an agent places a 40\,KB report into its context, it
does not pay for that report once; it pays on every subsequent turn of
that agent's life, because the whole history rides along each time. Now
add a second agent. An orchestrator receives a report and forwards it to
a writer subagent; the report now occupies \emph{two} context windows,
each re-billing it per turn. Add a fact-checker: three. This is not a
defect of any one framework---it is the default shape of the ecosystem.
LangGraph's handoff tool ``by default passes the full message
history''~\citep{langgraph}; Anthropic, engineering their own multi-agent
research system, measured agents at ${\sim}4\times$ the tokens of chat and
multi-agent systems at ${\sim}15\times$, attributing the overhead to
``duplicating context across agents'' and writing the sentence this work
is built on: \emph{``each handoff loses context''}~\citep{anthropic2025}.
The MAST failure taxonomy attributes \textbf{36.9\%} of multi-agent
failures to inter-agent misalignment---agents acting on divergent, stale,
or partial copies of what should have been the same
information~\citep{cemri2025}.

Copies are the root defect. A copy has no identity, no version, no owner,
no lifecycle. The moment the author corrects the report, every pasted copy
is silently wrong, and no mechanism exists to find them. The token cost is
what everyone feels; the missing lineage and the impossibility of
correction are what actually break swarms.

\paragraph{Thesis.} Passing work between computational actors admits
exactly three disciplines. \emph{Copy semantics} (message passing) sends
the bytes---simple, and every pathology above follows. \emph{Place
semantics} (shared memory) has both parties touch one location---this
fixes duplication but demands shared infrastructure and says nothing about
who may see what. \emph{Name semantics} sends a small, immutable,
attributed \emph{claim} and lets resolution be computed per consumer, at
the data, on demand. Distributed systems learned to prefer the third
discipline for exactly the reasons agent systems are now
rediscovering. This paper argues that the agent handoff is a
distributed-systems problem, that name semantics is its right shape, and
that a substrate built on names makes a handoff behave identically whether
its two ends sit in one process or on two continents. We substantiate the
argument with a working system, \textsc{waggle}~\citep{waggle2026}, and
its measurements.

\paragraph{Contributions.}
\begin{itemize}
  \item A framing of agent handoffs as a distributed-systems problem, with
    a four-boundary analysis (\S\ref{sec:problem}) showing that every
    boundary loses lineage and cannot propagate corrections.
  \item A substrate for \emph{name semantics} (\S\ref{sec:design}): a
    30-byte attributed token, a three-zone signed manifest, a sealed
    context-adaptive matcher (Alg.~\ref{alg:match}), and an append-only
    payload-free log with an exact-reconstruction guarantee.
  \item \emph{Receipts} (\S\ref{sec:receipts}): consumption contracts and
    a coverage fold (Alg.~\ref{alg:coverage}) that make delegated
    consumption verifiable without trusting the consumer's self-report.
  \item A \emph{structural lens} for source code (\S\ref{sec:lens})
    computed once at mint, so no parser ever runs on a serving path---%
    including the network edge.
  \item The \emph{three-radii} result (\S\ref{sec:radii}): the identical
    verb loop inter-process, inter-harness, and inter-machine.
  \item An evaluation (\S\ref{sec:eval}) of microbenchmarks and a live
    delegation, an honest comparison against lexical search, and a frank
    account of limits (\S\ref{sec:limits}).
\end{itemize}

% ===========================================================================
\section{The Handoff Problem}
\label{sec:problem}
% ===========================================================================

``Passing a resource between agents'' is four different problems wearing
one name, distinguished by which boundary the resource crosses.
Table~\ref{tab:boundaries} reads, column by column, as a ledger of what is
lost. The two rightmost columns are empty top to bottom: no boundary today
carries lineage, and none propagates a correction. That emptiness is the
problem this paper attacks.

\begin{table*}[t]
\centering
\small
\caption{The four boundaries of an agent handoff. Read the last two
columns top to bottom: they are empty. Lineage and correction do not
survive any boundary today.}
\label{tab:boundaries}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}p{2.9cm}p{4.1cm}p{4.1cm}p{2.4cm}p{2.0cm}@{}}
\toprule
\textbf{Boundary} & \textbf{How it works today} & \textbf{What travels} &
\textbf{Lineage} & \textbf{Corrections}\\
\midrule
Same harness, same model (orchestrator $\to$ subagent) &
Final message returns to the orchestrator, which pastes what the next
subagent needs; or shared filesystem paths by convention &
The full text again, per recipient; a path travels cheaply but carries no
version, access story, or receipt &
None---prose memory, gone at compaction &
Manual re-paste; stale copies persist silently\\
Same harness, mixed models (router / cascade) &
The same paste, now crossing a billing boundary &
The full text, re-tokenized under a different tokenizer &
None &
None\\
Across harnesses, one machine &
Files on disk plus convention files, or a human copy-pastes between panes &
A path with no semantics, or the text itself &
None---the filesystem records mtime, not meaning &
None---delete the file and one side breaks silently\\
Across machines / orgs &
Chat, email, tickets, shared drives; artifact-by-URL at the frontier &
Whole artifacts, or URLs whose resolution semantics no standard defines &
Whatever the messaging tool retains, divorced from the artifact &
Effectively impossible\\
\bottomrule
\end{tabular}
\end{table*}

\paragraph{Why provider optimizations do not save you.} Every vendor is
attacking the token cost inside its own walls: prompt caching discounts
byte-identical prefixes within a short TTL and a single account; harness
compaction summarizes a long context locally and lossily. These are real
savings. But each is scoped to one vendor's billing and serving boundary.
A cached prefix at one provider means nothing at another's tokenizer; a
compaction summary in one harness does not exist in another. Cross any row
of Table~\ref{tab:boundaries} and you pay full freight again, because what
crossed was a copy of bytes, and bytes carry no identity a foreign system
could recognize. This is the structural case for name semantics:
portability cannot be an optimization applied to copies. It must be a
property of the reference. A 30-byte token is the same 30 bytes in every
context window, every tokenizer, every vendor, every machine; what varies
is what it \emph{resolves to}, computed fresh, per consumer, where the
bytes live.

% ===========================================================================
\section{What the Colony Knew}
\label{sec:bees}
% ===========================================================================

A honeybee returns from a field two kilometres out carrying information
worth sharing: where, how far, how good. She does not carry the field
home, nor enough nectar for the colony to evaluate. She
\emph{dances}~\citep{vonfrisch1967}. The waggle run's angle against
vertical encodes direction relative to the sun; its duration encodes
distance; its vigour encodes quality. Then the colony does something every
distributed-systems engineer should study: each follower resolves the
reference \emph{herself}---flies her own flight, with her own senses, from
her own position. The dance is not the nectar; it is an attributed,
resolvable claim that the nectar exists. Recruitment is measurable, one can
count who followed and who returned to dance in turn, and the information
expires honestly: bees dance only while the source still pays, so no bee
chases stale copies of yesterday's directions, because there were never
any copies to chase.

This is \emph{stigmergy}: coordination through durable marks rather than
direct messages~\citep{grasse1959,theraulaz1999}. The mapping to a handoff
substrate is structural, not decorative (Table~\ref{tab:mapping}). The
colony solved the multi-agent handoff problem with zero shared context
windows: share names, not payloads; let each consumer resolve per its own
capability; make consumption observable; let stale claims die at the
source.

\begin{table}[t]
\centering
\small
\caption{The dance and the substrate. The correspondence is structural.}
\label{tab:mapping}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{3.7cm}p{4.0cm}@{}}
\toprule
\textbf{The dance} & \textbf{The substrate}\\
\midrule
Figure-eight encodes a vector in seconds & 30-byte token names an artifact plus its attribution\\
Each follower flies her own flight & Each consumer resolves \emph{its} projection (sealed matcher)\\
The follower's senses at the field & \tok{read}/\tok{search}: interrogate content on arrival\\
Countable recruitment & The funnel: resolve $\to$ read $\to$ run, as receipts\\
Dancers recruiting dancers & Lineage: children minted under parents\\
Dancing stops when nectar stops & Leases, supersession, revocation\\
The dance floor & The append-only log every mark lands on\\
\bottomrule
\end{tabular}
\end{table}

% ===========================================================================
\section{Design: Name Semantics}
\label{sec:design}
% ===========================================================================

\subsection{The token and the manifest}
A \emph{token} is 4--23 characters of Bitcoin base58 (default 8 public, 16
for private capability URLs), generated by rejection sampling to avoid
modulo bias. It is the same 30 bytes in every tokenizer. Behind it stands
an \emph{attribution manifest} in three zones with three mutability rules,
a separation that is load-bearing:

\begin{description}[leftmargin=1.1em,style=nextline]
  \item[Immutable core] set at mint, never changed: schema, token, target,
    sharer, channel, mint time, snapshot metadata, parent, the
    content-addressed snapshot reference, variants, and---optionally---a
    consumption contract (\S\ref{sec:receipts}) and a symbol-outline
    reference (\S\ref{sec:lens}). Signatures cover exactly this zone.
  \item[Versioned mutable] compare-and-swap by version: expiry,
    revocation, supersession. Every lifecycle change cites the version it
    was decided against.
  \item[Cosmetic mutable] last-writer-wins: campaign and labels.
\end{description}

Because signatures (Ed25519 over the canonical core bytes) cover only the
immutable zone, lifecycle and cosmetic churn never invalidate a
signature---the three-zone design's payoff, and a direct descendant of
capability discipline~\citep{dennis1966,miller2006}. Content over 64\,KiB
rides as a content-addressed reference~\citep{merkle1987,benet2014} rather
than inline, so the manifest stays small and the bytes stay verifiable
wherever they replicate.

\subsection{The sealed matcher}
A manifest carries one or more \emph{variants}, each a projection of the
artifact for a class of consumers, guarded by a match expression over four
dimensions: model family, harness, modalities, and posture. Selection is
\emph{sealed}: the algorithm is fixed and exposes no hooks, so the trust
claim ``the same context always receives the same projection'' survives
(Alg.~\ref{alg:match}). This is the resolve-per-consumer discipline of the
dance made deterministic: one name, many truthful renderings---a
Claude-tuned digest, an image for a vision agent, a transcript for an
agent without ears, fail-closed instructions for CI.

\begin{algorithm}[t]
\small
\caption{Sealed variant selection. Total over any manifest (mint
guarantees exactly one catch-all).}
\label{alg:match}
\DontPrintSemicolon
\KwIn{variants $V$; resolver context $c$}
\KwOut{index of the selected variant}
$best \leftarrow \bot$;\quad $bestSpec \leftarrow -1$\;
\For{$i \leftarrow 0$ \KwTo $|V|-1$}{
  \If{$\neg\, \textsc{Accepts}(V[i].expr, c)$}{\textbf{continue}\;}
  $s \leftarrow \textsc{Specificity}(V[i].expr)$
    \Comment*[r]{constrained dims, $0..4$}
  \If{$s > bestSpec$}{
    $best \leftarrow i$;\quad $bestSpec \leftarrow s$
      \Comment*[r]{ties: earlier declaration wins}
  }
}
\Return $best$\;
\BlankLine
\textbf{\textsc{Accepts}}$(e, c)$: every constrained dimension of $e$
admits $c$; an undeclared context value \emph{fails} a constrained
dimension; modalities match by superset.\;
\end{algorithm}

\subsection{The extraction boundary: determinism, not format}
\label{sec:opaque}

A PDF is a container. So is an MP4, a WAV, a scanned page. Somewhere inside is
something a consumer needs, and the tempting move is to reach in and transcode
everything to a clean string. The line the substrate actually draws is not
``text versus binary''---it is \emph{deterministic versus model-requiring}, and it
runs straight through the binary formats.

A PDF's embedded text layer and an HTML document's text are \emph{pure functions of
the bytes}: \texttt{pdftotext} reads a stream the file literally stores, stripping
tags reads the DOM, same bytes to same text with no model in the loop. These the
substrate \emph{does} extract, at mint, recording the extractor's identity and
signing the result into the core, so a PDF becomes searchable text that travels
\emph{with} the token and joins the lens stack of \S\ref{sec:lenses}. A scanned page,
audio, or video carries no text layer; recovering its content takes a \emph{model},
and there the substrate refuses. It pins the raw bytes and the projection says so:
\emph{``content is \texttt{audio/mp4}---the substrate does not read this; fetch the
bytes and interpret them with your own speech model.''}

Three reasons put the line exactly at determinism. \textbf{Competence}: the models on
the far side already see and hear, better than any pipeline we could ship, and the
gap widens with every generation while ours stands still---to OCR a page \emph{for}
such a model is to demote it. \textbf{Timing}: a model transcode is a projection
chosen in advance, at mint, by a substrate that does not know the question; the
consumer's own perception is a projection chosen \emph{at the point of use}, by the
party that does. A deterministic text-layer extraction escapes this critique
precisely because it is not a projection at all---it recovers the one text the format
already committed to, and discards nothing the layout did not. \textbf{The receipt}:
a substrate that runs a model must form a non-deterministic \emph{opinion} about the
bytes, and every receipt downstream inherits the drift---which is exactly what
invariants~I-1 (payload-free log) and~I-2 (sealed, reproducible matcher) forbid. A
deterministic extraction breaks neither: the log still records only that a token was
minted, and the matcher over the extracted text reproduces exactly. The receipts are
trustworthy \emph{because} the extractor cannot hallucinate. The signed extraction
therefore carries a \texttt{deterministic} flag---\texttt{true} for a text layer,
and \texttt{false} for a registered model extractor (opt-in, never a default), so a
reader weighs a coverage receipt over the extraction at its true worth.

The corrected claim is sharper than the one an earlier draft made. It is
\emph{not} that a path ``scores zero'' on a PDF---given \texttt{pdftotext}, an agent
with a path reaches \textbf{72\%} (\S\ref{sec:tier3}). It is that the extraction must
travel \emph{with} the reference. A path leaves it as a loose file the next agent
must locate and the receipt cannot vouch for; a corrupted, stale, or substituted
extraction produces the same confident receipt. The token instead \emph{delivers} a
deterministic extraction whose provenance is signed---and does so, on the PDF task,
at \textbf{583 bytes served against pasting's 23{,}651}, more accurately than the
path and a fortieth of the copy. That is the difference between a name that points at
content and one that carries it.

\subsection{The log is the truth}
Everything downstream of a mint is an \emph{event} in an append-only log,
and every view---manifest tables, funnel counts, lineage---is a fold over
that log~\citep{kreps2013}. Two invariants make the log safe to replicate
anywhere. First, events are \emph{payload-free by construction}: the event
type is exactly $\langle \textit{token}, \textit{stage}, \textit{actor},
\textit{time}, \textit{seq}, \textit{variant?}, \textit{regions?}\rangle$,
and no payload field \emph{exists}, so the receipt trail can never become
the leak. The actor is a coarse class---kind, model family, harness
class---never a version or an instance identifier. Second, reconstruction
is exact: folding the log rebuilds byte-identical state regardless of
arrival order, and is immune to duplicate and shuffled records
(Fig.~\ref{fig:log}). Migration is therefore a \emph{stream}: export the
log, replay it elsewhere, and the destination \emph{is} the source---same
tokens, same history, same receipts. This is the property that makes
``across machines'' transport plus replication rather than new
architecture.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  rec/.style={draw=line,fill=pale,minimum width=1.05cm,minimum height=0.6cm,inner sep=1pt},
  view/.style={draw=line,rounded corners=2pt,minimum width=1.9cm,minimum height=0.55cm,inner sep=1pt},
  >={Stealth[length=1.6mm]}
]
% the log
\node[rec] (m) at (0,0) {mint};
\node[rec,right=1mm of m] (e1) {resolve};
\node[rec,right=1mm of e1] (e2) {read};
\node[rec,right=1mm of e2] (mu) {revoke};
\node[rec,right=1mm of mu] (e3) {run};
\node[left=2mm of m,text=stone] {\itshape log};
\draw[decorate,decoration={brace,amplitude=4pt},stone]
  ($(m.north west)+(0,1mm)$) -- ($(e3.north east)+(0,1mm)$)
  node[midway,above=2pt,text=ink]{append-only, payload-free};
% the fold
\node[view,below=1.0cm of m,xshift=0.4cm] (f1) {manifests};
\node[view,right=2mm of f1] (f2) {funnels};
\node[view,right=2mm of f2] (f3) {lineage};
\node[left=2mm of f1,text=stone] {\itshape fold};
\foreach \x in {m,e1,e2,mu,e3}{
  \draw[->,stone] (\x.south) -- ($(f2.north)+(0,0)$);
}
\node[below=0.5cm of f2,text=stone,align=center]
  {\scriptsize shuffle-immune $\cdot$ duplicate-immune $\cdot$ byte-exact};
\end{tikzpicture}
\caption{The log is the truth; every view is a fold. Reconstruction is
order-insensitive, so the log replicates freely and migration is a
replay.}
\label{fig:log}
\end{figure}

\subsection{Interrogation, not transmission}
The consumer's first move is never a blind fetch. Holding a token affords
three instruments, all under one hard byte-budget contract (default
4\,KB): \tok{query} slices the \emph{metadata} by pointer; \tok{read}
slices the \emph{content}---a line window, a named section, a code symbol,
a JSON path; \tok{search} greps the content and returns matches with full
counts even when the list is truncated. Every response names the bytes it
spared you, and carries executable ``next'' steps so an agent that only
follows guidance still reaches every leaf. Retrieval becomes
interrogation: a consumer that needs three facts from a sixty-page report
ingests a few hundred bytes, ever---the follower's senses at the field,
not the field carried home. This is a direct instance of the end-to-end
argument~\citep{saltzer1984}: the substrate moves the question to the data
rather than the data to the question, and the marginal-value calculus of
when to stop reading is the forager's~\citep{charnov1976,pirolli1999}.

% ===========================================================================
\section{Receipts: Verification Without Trust}
\label{sec:receipts}
% ===========================================================================

The orchestrator's real question about a delegation is not ``what did the
subagent say'' but ``did it actually read what I gave it.'' Under copy
semantics this is unanswerable; under name semantics it is a query. Every
resolve, read, and search is an event, so the \emph{funnel}---stage counts
per token, with actor classes, never payloads---reports which handoffs were
consumed, which stalled, and which delivered.

Two mechanisms sharpen a raw count into a verdict. A \emph{consumption
contract}, declared at mint and signed into the immutable core, names the
regions a consumer must reach (up to eight line ranges, or a threshold
fraction of them). At serve time, when a \tok{read} window or a
\tok{search} hit overlaps a required region, the serving path stamps a
\emph{manifest-referencing bitmask}---an index into the signed
declaration, never bytes---onto the event, preserving the payload-free
invariant exactly as the variant field does. A \emph{coverage} fold then
$\mathrm{OR}$s these bitmasks and reports met/unmet with the untouched
regions \emph{named} (Alg.~\ref{alg:coverage}); because $\mathrm{OR}$ is
commutative and idempotent, the fold inherits the log's shuffle- and
duplicate-immunity for free. Finally, the judge's verdict rides as its own
stages---\tok{accepted}, \tok{rejected}---so the outcome is derivable from
counts (pending / accepted / rejected / contested) and a rejection's
response teaches the escalation: re-mint the same target under the same
parent for a stronger consumer, then supersede the rejected child, so the
escalation itself becomes lineage.

\begin{algorithm}[t]
\small
\caption{Contract coverage. A pure fold; misses are named, never
silent.}
\label{alg:coverage}
\DontPrintSemicolon
\KwIn{token $t$; its contract regions $R$ (up to 8); its events $E$}
\KwOut{$met$; the list of untouched regions}
$bits \leftarrow 0$\;
\ForEach{$e \in E$ \textbf{with} $e.token = t$}{
  \If{$e.regions \neq \bot$}{$bits \leftarrow bits \mathbin{|} e.regions$
    \Comment*[r]{OR: order- and dup-immune}}
}
$touched \leftarrow \textsc{PopCount}(bits)$\;
$permille \leftarrow \lfloor 1000 \cdot touched / |R| \rfloor$\;
$met \leftarrow permille \geq R.threshold$\;
$missed \leftarrow \{\, R[i] : \text{bit } i \text{ of } bits = 0 \,\}$\;
\Return $(met, missed)$\;
\end{algorithm}

This is precisely the telemetry loop that skill authors and model vendors
lack today. It is also, deliberately, a proof of \emph{consumption}, not
of comprehension---a distinction we return to in \S\ref{sec:limits}. The
signal it produces is the raw material for post-hoc, evidence-based model
routing, a complement to the pre-hoc, prompt-feature routers of current
practice~\citep{ong2024}, and its interrogation traces are the kind of
step-level process-supervision data that improves
reasoners~\citep{lightman2023,jin2025} and lets successful trajectories be
distilled into scaffolds for weaker consumers~\citep{wang2024awm}.

% ===========================================================================
\section{The Lens Stack}
\label{sec:lenses}
% ===========================================================================

Interrogation needs a \emph{surface to address}. A line window and a regular
expression treat every artifact as an undifferentiated string; that is the floor,
and for a log file it is enough. But a consumer handed a token wants to ask the
question the artifact's own shape invites --- a section of a document, a field of a
config, a function of a program --- and the shape is a property of the
\emph{resource type}, not of the consumer. So the substrate selects an addressing
surface --- a set of \emph{lenses} --- from the content type, and \tok{read} with no
address returns an overview that \emph{lists} the lenses the artifact affords. The
consumer discovers what it can ask; it never guesses.

This is a different axis from the \emph{variant} of \S\ref{sec:design}. A variant is
chosen by the \emph{consumer's} context --- a vision model and a text agent resolve
the same token differently. A lens is chosen by the \emph{artifact's} type --- and
every consumer of a markdown token can address it by section. Two projections,
orthogonal: one over who is asking, one over what is being asked of.

\begin{table}[t]
\centering\small
\caption{Lenses by resource type. Every text artifact gets the floor
(\texttt{lines}, \texttt{search}); the type \emph{adds} to it. Opaque formats join
the stack through extraction (\S\ref{sec:opaque}) --- a PDF becomes text and inherits
the text lenses --- while media the substrate will not read deterministically gets no
lens at all, by design: its bytes are served and the consumer's own model perceives
them.}
\label{tab:lenses}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}l l l@{}}
\toprule
\textbf{Resource type} & \textbf{Adds} & \textbf{Address by}\\
\midrule
any text & --- & \texttt{lines:A-B}, regex\\
\texttt{text/markdown} & \texttt{outline}, \texttt{section} & \texttt{section:"Pricing"}\\
\texttt{application/json} & \texttt{path} & \texttt{path:/deps/react}\\
source code & \texttt{symbol} & \texttt{symbol:resolve}\\
\midrule
PDF, HTML & \emph{via extraction} & as extracted text\\
audio, video, image & \emph{none} & fetch \& perceive\\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:lenses} is the whole dispatch. It is deliberately shallow: a lens is
an \emph{addressing} surface, never an interpretation, so adding one is adding a way
to name a region, not a way to read meaning --- which is why the same discipline that
governs a markdown section governs a code symbol, and why the extraction tier can
join the stack without a new invariant. The rest of this section is the deepest lens
in the table, worked in full.

\section{A Structural Lens for Code}
\label{sec:lens}
% ===========================================================================

Source code is the handoff the substrate must serve well, because it is
what orchestrators most often delegate. Text tools---line windows and
regular-expression search---treat a program as a string. An agent handed a
code token instead wants to know \emph{what the file is} before it greps:
its functions, methods, and types, with line ranges. We compute that
structure once, at mint, where the artifact is at hand, using an
incremental parser~\citep{treesitter} and its tag queries, and store the
result---a flat, document-ordered outline of definitions---as a
content-addressed blob beside the snapshot.

The design decision that matters is \emph{where parsing happens}. It
happens only at mint, on the machine that owns the bytes. Serving the
outline is a blob fetch and a budget-fitted render---no parser on any
serving path, and in particular none at the network edge, whose runtime
could not host a native parser in any case. The outline is thus authored
content derived from pinned bytes: a pure function of the snapshot and an
extractor version, so it is minted once and never invalidated, and
identical bytes deduplicate for free. A \tok{read} by symbol name resolves
to the definition's exact line range through the ordinary windowed-read
path, so contract stamping applies unchanged; a \tok{require symbol:$N$}
requirement resolves the symbol to a line range at mint and enters the
signed contract. The lens does for code what section headings do for
prose, uniformly across languages, without a single new invariant.
Extraction is measured at 2.3\,ms per thousand lines
(\S\ref{sec:eval})---amortized entirely into the one-time mint.

% ===========================================================================
\section{One Loop at Three Radii}
\label{sec:radii}
% ===========================================================================

The systems claim of the paper is that a handoff behaves identically
regardless of the distance between its two ends. A single daemon owns the
local store; every harness on a machine speaks to it over a
filesystem-permissioned domain socket, so cross-harness handoff on one box
is not a feature but the default. Daemons federate to one another over
authenticated transport; the same tokens graduate to a network edge by
replaying the log, with content-addressed snapshots replicated so that a
\tok{search} runs \emph{at the edge} against content whose source file
never left the author's laptop (Fig.~\ref{fig:radii}). Because the shim
between a harness and the daemon adds no semantics---it pumps
JSON-RPC frames between a pipe and a socket---``across machines'' is
transport plus replication, not new architecture. An agent's loop---mint,
hand off one line, resolve, interrogate, report---is byte-for-byte the same
at all three radii. That invariance is the dividend of name semantics, and
it is why we claim distributed subagents are the near future rather than a
special case: the programming model does not change as the swarm spreads
from one core to many machines. The lineage here is old and deep---tuple
spaces made coordination a property of a shared
medium~\citep{gelernter1985}; named-data networking made the network route
by name rather than by host~\citep{jacobson2009}; leases made freshness a
bounded promise rather than a cache guess~\citep{gray1989}. \textsc{waggle}
is these ideas applied to the artifact a swarm of agents passes around.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  font=\footnotesize,
  >={Stealth[length=1.8mm]},
  proc/.style={draw=line,fill=pale,rounded corners=1pt,minimum width=1.5cm,minimum height=0.5cm,inner sep=1pt},
  dmn/.style={draw=line,thick,minimum width=1.45cm,minimum height=0.55cm,inner sep=2pt},
  radius/.style={draw=stone,dashed,rounded corners=5pt,inner sep=7pt}
]
% --- radius 1: one machine (top) -----------------------------------------
\node[proc] (h1) at (0,1.05) {Claude Code};
\node[proc] (h2) at (0,0.35) {Codex};
\node[proc] (h3) at (0,-0.35) {Cursor};
\node[dmn] (d1) at (3.0,0.35) {\textsf{waggled}};
\foreach \h in {h1,h2,h3}{\draw[->] (\h.east) -- (d1.west);}
\begin{scope}[on background layer]
\node[radius,fit=(h1)(h3)(d1)] (r1) {};
\end{scope}
\node[text=stone,font=\scriptsize,anchor=west] at ($(r1.north west)+(0.1,0.22)$)
  {radius 1: one machine \ \textbullet\ \ shims add no semantics};
% --- radii 2--3: machines & network (bottom) -----------------------------
\node[dmn] (d2) at (3.0,-2.55) {\textsf{waggled}};
\node[dmn,fill=pale] (edge) at (0.35,-2.55) {edge};
\draw[<->] (d1.south) -- (d2.north)
  node[midway,fill=white,inner sep=1pt,text=stone,font=\scriptsize]{federation};
\draw[<->] (edge.east) -- (d2.west)
  node[midway,fill=white,inner sep=1pt,text=stone,font=\scriptsize]{log replay};
\begin{scope}[on background layer]
\node[radius,fit=(edge)(d2)] (r2) {};
\end{scope}
\node[text=stone,font=\scriptsize,anchor=west] at ($(r2.south west)+(0.1,-0.22)$)
  {radii 2--3: machines \& network \ \textbullet\ \ grep runs at the edge};
\end{tikzpicture}
\caption{One token at three radii. The verb loop is identical; only the
transport lengthens. Computation travels to the data, so the edge answers
\tok{search} against snapshots whose source files never left the author's
machine.}
\label{fig:radii}
\end{figure}

Consumption is protocol-shaped rather than SDK-shaped: the substrate is a
Model Context Protocol server~\citep{mcp2024}, one configuration line in
any MCP-speaking harness. The interrogation verbs are exposed as
\emph{tools} because the protocol assigns model-controlled actions to
tools; the passive faces of a token---enumeration, plain read, and
subscription to lifecycle updates---are exposed as \emph{resources}, so a
correction propagates to holders as a protocol-native push. The frontier's
agent-to-agent efforts standardize artifact-by-URL and stop
there~\citep{a2a2025}; name semantics is the resolution discipline that
gap wants.

% ===========================================================================
\section{Evaluation}
\label{sec:eval}
% ===========================================================================

We evaluate three questions: is the reference cheap enough to prefer over
a copy; does verification-without-trust work on a real delegation; and how
does interrogation-through-a-token compare with the lexical search agents
use today.

\subsection{Microbenchmarks}
Table~\ref{tab:perf} reports the hot paths. Measurements are taken on
commodity hardware with a warm store; the domain-socket figure is
end-to-end through the shim, daemon dispatch, sealed match, event append,
and reply. The cache-hit resolution at 39\,ns and the million-event fold
at 334\,\textmu{}s establish that neither resolution nor the receipt
machinery is a bottleneck; the 323\,\textmu{}s socket round-trip is
dominated by the durable append's \texttt{fsync}, which is the price of the
$C\!$-1 durability guarantee, not of the abstraction. Symbol extraction at
2.3\,ms/kLOC is paid once, at mint.

\begin{table}[t]
\centering
\small
\caption{Hot-path measurements. The abstraction is not the cost; durable
persistence is.}
\label{tab:perf}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Operation} & \textbf{Latency}\\
\midrule
Cache-hit resolution (sealed match + stamp) & 39\,ns\\
End-to-end resolve over domain socket (p50) & 323\,\textmu{}s\\
Durable event append (real \texttt{fsync}) & 39\,\textmu{}s\\
Million-event funnel fold & 334\,\textmu{}s\\
Symbol extraction (per 1000 lines, one core) & 2.3\,ms\\
Outline render under 4\,KB budget & ${\sim}0.1$\,ms\\
Edge resolve, full HTTP--worker path (p50) & 1.2\,ms\\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The cost, measured.} Table~\ref{tab:cost} prices the same
delegation three ways against a deliberately strong baseline---a copy that
enjoys within-vendor prompt caching---and Figure~\ref{fig:cost} sweeps
artifact size at the five-consumer, five-turn operating point. Two
properties keep the account honest. The baseline is \emph{cached}, not the
straw-man of naïve re-sending; and the reported ratio is
\emph{tokenizer-invariant}---the tokenizer cancels in a ratio---so the
headline does not turn on a choice of BPE. The figure carries one central
fact: waggle's cost is \emph{flat in artifact size}, because the 30-byte
line and the bounded projection do not grow with the report, while the copy
grows linearly; the ratio therefore widens without bound, from
$2\times$ on a single-shot 4\,KB hand-off, through $44\times$ at the
40\,KB/five-consumer/five-turn cell, to $329\times$ under deep delegation.
We are equally plain about the floor: at $H{=}T{=}1$ the advantage is only
$2\times$, and the substrate earns its keep when an artifact is
\emph{shared or revisited}, not on one blind read. The deeper point is in
the columns a copy does not have. Under name semantics the author
\emph{knows} the report was resolved and read, the correction reached the
holder that resolved early, and the exchange replays on any machine; under
copy semantics that knowledge is not expensive---it is nonexistent. Every
number here is regenerated from source by the benchmark harness
(\texttt{just bench-paper}); the pre-registered cost model and its threats
to validity are in the accompanying design note.

\input{generated/cost_table.tex}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.86\columnwidth, height=5cm,
  xlabel={artifact size (KiB)}, ylabel={cost (tokens)},
  xmode=log, ymode=log, log ticks with fixed point,
  xtick={4,16,40,160,640}, xticklabels={4,16,40,160,640},
  legend pos=north west, legend cell align=left,
  legend style={font=\scriptsize, draw=none, fill=none},
  grid=both, major grid style={black!12}, minor grid style={black!5},
  tick label style={font=\footnotesize}, label style={font=\small},
  every axis plot/.append style={very thick},
]
\addplot table[x=s_kib, y=copy_naive]  {generated/cost_sweep.dat};
\addplot table[x=s_kib, y=copy_cached] {generated/cost_sweep.dat};
\addplot table[x=s_kib, y=waggle]      {generated/cost_sweep.dat};
\legend{copy (no cache), copy (cached), waggle}
\end{axis}
\end{tikzpicture}
\caption{Handoff cost versus artifact size at $H{=}5$, $T{=}5$, $R{=}1$.
waggle is flat---it never sends the artifact---while the copy grows
linearly; the ratio is tokenizer-invariant. Regenerated from source by
the benchmark harness.}
\label{fig:cost}
\end{figure}

\subsection{A live delegation}
\label{sec:live}
We ran the loop against this system's own source repository. An
orchestrator minted a source file with a snapshot and a symbol contract
requiring the definition of one function, then handed a freshly spawned
subagent nothing but the one-line reference and a question whose answer lay
inside the required definition. The subagent was instructed to work only
through the substrate. Its interrogation was, in order: resolve; a single
symbol read (which returned the definition's exact line range without any
window guessing); two searches; and a report of completion. The receipts
recorded this independently: the coverage fold flipped from unmet---with
the required region \emph{named}---to met at full coverage, and the funnel
read $\langle\textit{resolve}{:}1, \textit{read}{:}\text{(orchestrator's
overview}+\text{subagent's reads)}, \textit{run}{:}1\rangle$, whose
arithmetic matched the subagent's self-reported command list exactly. The
orchestrator did not have to trust the report; the coverage flip was
proof. We also confirmed the cross-connection path: a client subscribed to
a token on one connection received a lifecycle-update notification when a
\emph{different} connection revoked it. This is a single trial under
favorable conditions (a capable model, explicit instructions); we present
it as an existence proof of the mechanism, not as a controlled study, and
we say so plainly.

\subsection{Receipt reliability: a controlled measurement}
\label{sec:tier2}
\input{generated/tier2.tex}
The live trial shows the mechanism \emph{fires}; it does not say how
\emph{reliably}. To characterise the detector we run a pre-registered
controlled experiment (design note~\S3): $\benchTwoTrials$ trials per
condition against a $\benchTwoRegions$-region consumption contract, a
quarter of the subagents \emph{bluffers}---modelled to report completion
without reading. Every trial's interrogation routes through the real
coverage machinery---region-touch events, folded and judged against the
contract, exactly as in the live run; only the agent's read pattern is a
model, and it lives behind a driver seam so a live model substitutes for the
script without changing the measurement. The parameters are fixed in
advance: a genuine reader touches each required region with probability
$0.98$, a bluffer only incidentally ($0.04$), and through the side door a
genuine reader bypasses the substrate with probability $0.35$.

Table~\ref{tab:tier2} reports the outcome and Figure~\ref{fig:roc} the
coverage ROC. Two findings matter. Precision is essentially perfect in both
conditions---the coverage signal almost never mistakes a bluffer for a
genuine consumer, so a \emph{passing} receipt is trustworthy---and the
graded signal separates the two populations cleanly (ROC AUC
$\benchTwoAuc$). The second finding is the one the mechanism lives or dies
on: under seal the receipts miss only $\benchTwoSealedFnr$ of genuine
consumers, but through the side door that false-negative rate rises to
$\benchTwoSideFnr$. The seal is not cosmetic; it is what converts a
suggestive signal into a dependable one. Substituting live agents for the
scripted read pattern is the pre-registered next step, and the detector
properties measured here are precisely what that step would put to the test.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=0.86\columnwidth, height=5cm,
  xlabel={false positive rate}, ylabel={true positive rate},
  xmin=0, xmax=1, ymin=0, ymax=1.02,
  legend pos=south east, legend cell align=left,
  legend style={font=\scriptsize, draw=none, fill=none},
  grid=both, major grid style={black!12}, minor grid style={black!5},
  tick label style={font=\footnotesize}, label style={font=\small},
]
\addplot[very thick] table[x=fpr, y=tpr] {generated/tier2_roc.dat};
\addplot[dashed, black!45] coordinates {(0,0) (1,1)};
\legend{coverage signal, chance}
\end{axis}
\end{tikzpicture}
\caption{Coverage ROC over all trials: the signal a passing receipt gives
against ground-truth consumption. Regenerated from source by the benchmark
harness.}
\label{fig:roc}
\end{figure}

\subsection{The substrate under load}
\label{sec:tier3}
\input{generated/tier3_gate.tex}
The measurements above test the receipt machinery. This one tests the
\emph{handoff}, against the competitor the introduction named. Twelve artifact
shapes---\texttt{text}, \texttt{markdown}, \texttt{code}, \texttt{pdf},
\texttt{voice}, \texttt{video}, and the shapes real work has: a \emph{folder} of
twelve runbooks, a \emph{$>$2000-line} source file, a \emph{multi-hop} pointer
chain, a \emph{reasoning} task whose answer no single line contains, and a
\emph{big tree} of 180 files across 310\,KB, interrogated twice---once for a
needle, once for a \emph{count}. Nine models across two families answer the same
question under four \emph{turn-matched} arms. \emph{copy} pastes the artifact;
\emph{reference} hands a filesystem path plus \tok{ls}, \tok{grep} and
\tok{open}; \emph{waggle} hands a token; \emph{waggle+gate} adds the one move no
other arm can make. Of 1728 runs, 1704 are analysed; the rest are excluded for
unrecovered transport errors, and we say so rather than counting them as model
failures. Every arm is given the abilities a real agent has---the \emph{reference}
path arm can \tok{ls}, \tok{grep}, and run \tok{pdftotext} on a binary, so the
baseline is a steelman, not a straw man. Because every arm answers the same
(artifact, model) pairs, the arms are \emph{paired}, and we report the sign test on
the runs where two arms disagreed---the only runs that carry information about the
difference between them.

\paragraph{Parity with the paste, at a thirty-second of the context.} Across all
1704 runs, \emph{waggle} answers 96.5\,\% and \emph{copy} 96.7\,\%. Paired, they
are indistinguishable: 13 runs where waggle is right and copy wrong, 15 the other
way ($p=0.85$). \emph{waggle+gate} answers 98.6\,\% against copy's 96.7\,\%, and we
still decline to call that a win: 14 discordant pairs to 6, $p=0.12$. \textbf{The
honest claim is parity, not victory}---and parity is bought at \textbf{2{,}579 bytes
of ingested context against pasting's 81{,}666}, a factor of 32, for 10$\times$ less
money ({\$}4.45 against {\$}48.53 for the same 1704 answers). Pasting was never
inaccurate. It was expensive, and it left no record; the substrate's case is that
you can stop paying for it without paying for it in correctness.

\paragraph{Where the path falls short.} The \emph{reference} arm---a path and
ordinary file tools, which is what an agent is handed today---answers \textbf{90.4\,\%}
with a fair toolset, and the token's edge over it is real but modest:
\textbf{34 paired wins to 9} ($p<0.05$). We stress the fairness, because an earlier
draft of this work reported the path at zero on binary formats---a number that was
an artifact of \emph{our} crippled baseline, not of file paths, and we corrected it.
The token's advantage now lives in specific shapes rather than a blanket win. On
\texttt{pdf} the path reaches only \textbf{78\,\%} \emph{with} \texttt{pdftotext} in
hand, paying 8{,}694 bytes, because the tool dumps one flat blob and the needle sits
past the first window; the token, extracting the same PDF once at mint and serving a
lensed slice, answers \textbf{100\,\%} at \textbf{750 bytes} (\S\ref{sec:opaque}).
The extraction must travel \emph{with} the reference, deterministically and signed;
a path leaves it a loose file the receipt cannot vouch for. And on the multi-hop
chain the path collapses to \textbf{29\,\%}, for a second and more interesting
reason: \tok{grep} cannot search for a section whose name it has not read yet.
Lexical search is superb at finding what you can already describe; it has no answer
at all for structure you must \emph{traverse}. We remain equally clear about the
other direction: on a small local text file a \tok{grep} is still cheaper than a
token, and the substrate's case was never that it beats \tok{grep} at \tok{grep}'s
own game.

\paragraph{But my agents share a filesystem.} The sharpest objection to the whole
premise, and the \emph{reference} arm is precisely it: subagents on one machine,
handed a path, free to \tok{ls}, \tok{grep}, \tok{open}, and \tok{pdftotext}. A path
is not a copy---both agents point at the same inode---so a shared filesystem
\emph{is} pass-by-reference for the bytes, and the 90\,\% the arm scores says so.
The distinction the substrate rests on is not copy-versus-reference; it is that a
path is a \emph{location}, and a location cannot answer three questions a handoff
eventually must. \textbf{Was it read, and which parts?} \tok{cat} and \tok{grep}
leave no trace an orchestrator can query, so the receipt result of this section
(\S\ref{sec:tier2} and above)---99\,\% correct when the required regions were
consumed, 20\,\% when not---is not merely hard to reproduce with a filesystem; it is
\emph{impossible in principle}, because reading a file records nothing. \textbf{Which
version?} A path names mutable bytes: correct the file mid-task and one agent read
the old copy, another the new, with no lineage to tell them apart---the divergent-copy
failure the introduction quantified, occurring \emph{with} a shared filesystem, not
only with pasting. \textbf{Reachable from where?} \texttt{file:///…} means nothing to
an agent in another container or at the network edge, while the same token resolves
unchanged (\S\ref{sec:radii}). Where none of these bind---one machine, a short task,
a disciplined agent doing ranged reads, and nothing to audit---a path is the right
tool and the substrate is overhead, exactly as the fourth limitation of
\S\ref{sec:limits} states. The case for name semantics is accountability,
versioning, and reach; it was never that the filesystem duplicates bytes.

\paragraph{Where the paste collapses.} It collapses exactly once, and the shape
is worth stating plainly, because it is the one place the ceiling is not a
ceiling. Asked how many files in a 180-file tree mention a deprecated call,
\emph{copy}---holding all 310\,KB in its context---answers correctly
\textbf{61\,\%} of the time. \emph{waggle+gate} answers \textbf{100\,\%}. The
paste has every byte and cannot count them; a projection that returns the seven
matching files, and only those, can. Pasting wins where an artifact fits in a
window and the question is a lookup. It loses where the question is an
\emph{aggregate over structure}, and no amount of context repairs that: the
context was never the missing thing.

\paragraph{The receipt as an oracle.} The strongest result in the sweep is not an
arm's score. Of the 754 runs whose coverage receipt showed the author's required
regions \emph{consumed}, \textbf{99\,\% were correct}. Of the 20 whose receipt
showed them \emph{unconsumed}, \textbf{20\,\%} were---a collapse of seventy-nine
points, visible to an orchestrator that never read the answer, never read the
region, and never graded anything. That is the property a path cannot have, a
copy cannot have, and no quantity of context can buy: not a cheaper handoff, but
an \emph{accountable} one. And read as a \emph{control}, the receipt earns its
keep: \emph{waggle+gate} beats plain \emph{waggle} \textbf{13 paired wins to 4}
($p<0.05$)---the one place the gate delivers a significant gain is exactly where an
agent would otherwise conclude from too little.

\paragraph{The gate helps where the failure is retrieval, and hurts where it is
reasoning.} \emph{waggle+gate} changes exactly one thing: before accepting an
answer, the orchestrator reads the coverage receipt and declines any claim the
consumer's own trail does not support---without reading the answer, and without
naming the region. Where an agent's failure mode is \emph{concluding from too
little}, it works: the tree count rises from 89\,\% to 97\,\%, the multi-hop
chain from 83\,\% to 88\,\%, the large source file from 89\,\% to 97\,\%. On the
\emph{reasoning} task it does the opposite, and we report it rather than bury it:
\textbf{84\,\% ungated falls to 78\,\% gated}. The mechanism is visible in the
traces. A model reads six of eleven runbooks, answers, and is refused; it reads
the rest, and \emph{changes its answer}---sometimes from right to wrong. The gate
cannot distinguish a consumer that concluded too early from one that had already
seen enough, because a receipt records what was \emph{served}, never what was
\emph{understood}. Forcing more evidence on a model whose difficulty is judgement
rather than retrieval gives it room to argue itself out of a correct answer. A
completeness contract is a check on retrieval, and should be written where
retrieval is the risk; \S\ref{sec:limits} returns to this.

\paragraph{The benchmark's first job was to find our bugs---and the second was to
stop needing a model to do it.} It found seven, and every one was ours. A
zero-match \tok{search} stamped \tok{read} on every file it had swept, so a
single fruitless grep took a folder's coverage from $0/11$ to $11/11$ and the
gate certified a tree nobody had opened: bytes touched by a \emph{matcher} and
content served to a \emph{consumer} are different ledgers, and only the second is
a receipt. A directory projection truncated at 25 of 180 files while reporting no
total and no cursor, which reads as a complete table of contents for a tree seen
at 14\,\%. A lens fanned out across a folder budgeted each file at
$\texttt{max\_bytes}/6$---a hardcoded six-file assumption---and so served nine of
eleven runbooks and silently dropped two, which made a \texttt{files:all} contract
\emph{unsatisfiable in one call by construction}: the one move that closes a
completeness contract could not close it, every consumer fell back to fetching
children one at a time, and the weakest of them exhausted their turns and
answered nothing at all while the gate, correctly and uselessly, refused to
believe them. And a fan-out that overran the caller's byte budget by 15\,\% was
worse still: the client showed its model the first 4{,}500 bytes and dropped the
rest, while our receipt certified all ten files as read---a receipt attesting to
bytes the consumer never received. \textbf{A byte budget is a contract with the
caller, and a receipt may only attest to what the consumer got.}

We found these one at a time, each costing a sweep, because we were using a
language model as a bug detector. Every one of them was deterministic. The
correction is a \emph{conformance suite}: a scripted consumer, no model, no
tokens, that drives the substrate and asserts these invariants in seconds. It
simulates two consumers, and both matter. The \textbf{oracle} plays perfectly,
using the affordances the substrate advertises in the order it advertises them;
if the oracle cannot close a contract in a handful of calls, no model will, and
\emph{the contract is a trap rather than a check}. The \textbf{lazy} consumer
answers having read nothing, and the gate must refuse it; a gate that passes this
is not a gate. Twenty-nine checks now gate the build. We report this because it
is the methodological lesson of the work and it generalises past this system: an
evaluation that looks only at answers will mark the model down and ship the
defect.

\subsection{Interrogation versus lexical search}
For a code artifact crossing a delegation boundary, the token's structural
layer changes the interaction. Locating and reading a function with a
lexical tool is typically three operations---a name search returning
definition and call sites, a first windowed read with a guessed extent,
and often a corrective second read---and leaves no record. Through the
token it is one operation, \tok{read} by symbol name, returning the exact
extent (pinned at mint, so stable against later edits), and stamping a
receipt. The token also works where a filesystem tool cannot exist---a
remote consumer with no local checkout---because the search runs where the
bytes live and the matches travel back. We are equally clear about where
lexical search remains superior: the unminted workspace, where an agent
greps across a whole checkout that nobody minted, is the lexical tool's
home and the substrate offers nothing there by design; the token's own
\tok{search} is lexical-parity, not lexical-superior, its advantage being
the structural layer around it; and structural \emph{queries}
(``every use of $f$ within a type'') remain future work.

% ===========================================================================
\section{Limitations and Outlook}
\label{sec:limits}
% ===========================================================================

We hold the mechanism to an honest account. First, receipts prove
\emph{consumption}, not \emph{comprehension}: an agent can read every
required line and still ignore it, and the \tok{run} stage is
self-reported. The served-byte and coverage signals are the hard evidence;
\tok{run} is corroboration. Second, on a shared machine a consumer can
bypass the substrate and read the file directly, producing a false
negative---the ``side door.'' The remedy is sealing: moving the source
into a vault so the token is the only access path, which the system
supports and which we recommend whenever receipts carry consequences; the
reliability of receipts under seal, across many trials, is the metric that
would settle the mechanism's value; \S\ref{sec:tier2} measures it, and the
side door roughly sextuples the false-negative rate. Measuring it against
\emph{live} agents rather than a modelled read pattern remains open.
Third, any behavioral signal used for routing invites Goodhart's law; we
keep interrogation-shape features as inputs to an outcome-labeled judgment
rather than as free-standing rewards, mirroring the finding that
intermediate rewards help little while outcome supervision
drives gains~\citep{jin2025}. Fourth, the substrate earns its keep only for
\emph{produced artifacts that cross a delegation boundary}; for
shared-workspace exploration the filesystem is already the reference and
the substrate should stay out of the way. Fifth---and \S\ref{sec:tier3} says it
in numbers---\textbf{the token does not beat the paste on accuracy; it ties
it}, 96.5\,\% against 96.7\,\%, 13 paired wins to 15. We report the gated arm's
98.6\,\% and decline to claim it over the paste: 14 discordant pairs to 6 is
$p=0.12$, short of significance. And against a \emph{fair} path---one that may
\texttt{ls}, \texttt{grep}, and run \texttt{pdftotext}---the token's win shrinks
from the decisive margin an earlier draft claimed to a real but modest 34 paired
wins to 9; the path, given proper tools, answers 90\,\%, not the zero our own
crippled baseline once reported. What the substrate buys, then, is not a blanket
accuracy win but \emph{cost, specific-shape wins, and accountability}---a
thirty-second of the context, a tenth of the money, a decisive edge on PDFs and
multi-hop structure, and a receipt---and it is worth saying that plainly rather
than dressing parity as a rout. Nor does the token always cost less: on a small
local text file a \tok{grep} is cheaper, and the substrate's case was never that
it beats \tok{grep} at \tok{grep}'s own game.

Sixth, and most uncomfortably: \textbf{the gate made one shape worse}. On the
reasoning task, refusing answers the receipt did not back drove accuracy
\emph{down}, from 94\,\% to 88\,\%. The traces show why. A model reads six of
eleven runbooks, answers correctly, is refused, reads the rest, and changes its
answer to a wrong one. A receipt records what was \emph{served}, never what was
\emph{understood}, so the gate cannot tell a consumer that concluded too early
from one that had already seen enough---and forcing more evidence on a model
whose difficulty is judgement rather than retrieval gives it room to argue
itself out of a correct answer. The lesson is not that completeness contracts
fail; it is that they are a check on \emph{retrieval} and belong where retrieval
is the risk. Where it is---counting across a large tree, following a multi-hop
chain---the same gate is worth eleven and eleven points, and over the whole sweep
it beats the ungated token \textbf{13 paired wins to 4} ($p<0.05$). An
orchestrator should write the contract it can defend, and not one it cannot.

Finally, our corpus plants a known needle in most artifacts: that makes
correctness and coverage exactly gradable, but it is largely a retrieval task,
and the two shapes that are not---the chain and the reasoning judgement---are
precisely where the arms diverge and where our own mechanism misfires. A harder
corpus is the obvious next measurement, and we expect it to be less flattering.

The outlook follows from the three-radii result. As harnesses routinely
fan work across machines and across vendors, the copy discipline's costs
compound superlinearly---more copies, more billing boundaries, more stale
state, more of the misalignment that already explains a third of
failures~\citep{cemri2025}. A substrate whose programming model is
invariant to distance is not a convenience at that scale; it is the
condition under which the scale is manageable at all. The colony ran a
swarm on shared names, per-consumer resolution, observable recruitment, and
honest expiry, with no shared context window anywhere. The argument of
this paper is that our swarms should be built the same way, and that the
sooner the handoff becomes a reference rather than a copy, the less of the
future we spend paying for photocopies.

% ===========================================================================
\section*{Acknowledgments}
The design corpus, specification, and implementation of \textsc{waggle}
are developed in the open~\citep{waggle2026}. Portions of the engineering
and prose were produced with AI assistance under the author's direction
and review.

\balance
{\small
\bibliographystyle{plainnat}
\bibliography{references}
}

\end{document}
